---
title: "Accounting consistency"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accounting consistency}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(panel990)
```

990 financial data carries **deterministic accounting identities**: a total is
the sum of its parts, a net equals gross minus expenses, and on the balance
sheet assets equal liabilities plus net assets. `panel990` ships these as a
registry and lets you **check** whether rows satisfy them and **reconcile** rows
that do not -- changing the values as little as possible.

## The identity registry

```{r}
data("accounting_identities", package = "panel990")
by_id <- unique(accounting_identities[, c("identity", "section", "type")])
table(by_id$section, by_id$type)
```

58 identities across revenue (Part VIII), functional expenses (Part IX), and the
balance sheet (Part X). Each is a linear combination of ef2 fields that must
equal zero; for example the contributions subtotal:

```{r}
accounting_identities[accounting_identities$identity == "rev_contributions_subtotal",
                      c("variable", "coefficient")]
```

## Checking

`accounting_check()` evaluates every identity whose fields are present and, by
default, returns only the violations. Start from a consistent revenue row:

```{r}
row <- data.frame(
  EIN2 = "A", TAX_YEAR = 2020,
  F9_08_REV_CONTR_FED_CAMP = 100, F9_08_REV_CONTR_MEMBSHIP_DUE = 50,
  F9_08_REV_CONTR_FUNDR_EVNT = 30, F9_08_REV_CONTR_RLTD_ORG = 20,
  F9_08_REV_CONTR_GOVT_GRANT = 200, F9_08_REV_CONTR_OTH = 100,
  F9_08_REV_CONTR_TOT = 500,
  F9_08_REV_OTH_INVEST_INCOME_RLTD = 10, F9_08_REV_OTH_INVEST_INCOME_UBIZ = 5,
  F9_08_REV_OTH_INVEST_INCOME_EXCL = 35, F9_08_REV_OTH_INVEST_INCOME_TOT = 50
)
accounting_check(row)                     # consistent -> no violations
```

Now imagine two contribution components and one investment column were imputed
independently and no longer add up:

```{r}
bad <- row
bad$F9_08_REV_CONTR_GOVT_GRANT <- 180     # true 200
bad$F9_08_REV_CONTR_OTH <- 90             # true 100
bad$F9_08_REV_OTH_INVEST_INCOME_EXCL <- 30  # true 35
accounting_check(bad)
```

The check names the broken identities and the exact gap: contributions short by
30, the investment column short by 5.

## Reconciling

`reconcile()` adjusts the values as little as possible (weighted least squares)
to satisfy the identities. Because the org **reported** its totals, we hold
those fixed and let the imputed components move:

```{r}
imputed <- c("F9_08_REV_CONTR_GOVT_GRANT", "F9_08_REV_CONTR_OTH",
             "F9_08_REV_OTH_INVEST_INCOME_EXCL")
fixed <- setdiff(names(bad), c(imputed, "EIN2", "TAX_YEAR"))

rec <- reconcile(bad, fixed = fixed)
rec[, c(imputed, "F9_08_REV_CONTR_TOT", "F9_08_REV_OTH_INVEST_INCOME_TOT")]
accounting_check(rec)                     # identities restored
```

The 30 contributions gap splits evenly across the two free components
(195 / 105); the single free investment column absorbs its 5 (35). Every
identity now holds and the reported totals are untouched. Use `weights=` to make
some values more resistant to change than others (e.g. by imputation
uncertainty), or a different `fixed=` set to hold different values.

## When does this matter?

A useful subtlety: **whole-row linear imputation preserves linear identities.**
If the years bracketing a gap both balance, then interpolating (or averaging, or
carrying forward) the *entire* row also balances -- it is an affine combination
of balanced vectors. So identities break mainly under:

- **partial missingness** -- some cells observed, some imputed (the case above);
- **rounding** -- `as_integers` rounds each cell independently, so sums drift;
- **filer error** in the source data -- which `accounting_check()` surfaces on
  observed rows as a data-quality signal, not just an imputation artifact.

## Scope and caveats

The registry is a curated, high-confidence set: column splits, subtotals,
net-of-expense lines, the revenue grand total, and the balance-sheet equation.
It deliberately omits identities that cannot close cleanly from the 1x1 tables
alone -- the expense line-24 write-ins live in a one-to-many detail table, and
total assets / net-asset composition hit form-version variants and
donor-restriction-vs-fund-accounting differences. Memo lines ("of which"
amounts) are inequalities, not sums, and are excluded. All identities are scoped
to the full 990 (`form_scope = "PC"`); the 990EZ has its own, smaller set.
```
