---
title: "Balancing a panel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Balancing a panel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(panel990)
```

`panel_balance()` trims a panel to a **balanced rectangle** -- a block of years
in which every retained organization is observed in every retained year. It is
the complement of `panel_complete()`: completion *fills* gaps to reach a full
rectangle, balancing *drops* organizations and years to reach one.

```{r}
df <- data.frame(
  EIN2     = c(rep("A", 4), rep("B", 4), rep("C", 3), rep("D", 2)),
  TAX_YEAR = c(2019:2022,   2019:2022,   2019:2021,   c(2019, 2020))
)
```

`A` and `B` are observed all four years; `C` misses 2022; `D` only appears in
2019--2020.

## The year window is the trade-off

A **wider** window keeps more years but fewer organizations; a **narrower** one
keeps more organizations. By default `panel_balance()` uses the full observed
range -- the strictest balance:

```{r}
full <- panel_balance(df)
unique(full$EIN2)          # only orgs present in all of 2019-2022
attr(full, "balance")
```

Restrict the window and more organizations qualify:

```{r}
win <- panel_balance(df, years = 2019:2021)
unique(win$EIN2)           # C now qualifies; D still does not
```

## Let it find the largest block

If you would rather have the package pick the window that retains the most data
(organizations x years), use `strategy = "max_rectangle"`:

```{r}
best <- panel_balance(df, strategy = "max_rectangle")
attr(best, "balance")[c("years", "orgs_kept")]
```

Here the 3-year x 3-organization block (2019--2021, A/B/C = 9 observations)
beats the full 4-year x 2-organization block (8), so that is what it returns.
`min_years` sets a floor on how short a balanced window may be.

## Notes

The `"balance"` attribute always records the retained years and the
keep/drop counts. If no organization spans the window, the result is simply
empty (zero rows) rather than an error -- a useful signal that your window is
too demanding. And because balancing only *selects* rows, run `deduplicate()`
first if an organization can have more than one filing per year, so each cell of
the rectangle is a single record.
