---
title: "1. Downloading and assembling efile tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Downloading and assembling efile tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(panel990)
```

`panel990` retrieves IRS 990 efile tables from the NCCS S3 bucket, merges the
tables that describe a filing, and stacks years into one panel. This tutorial
specifies everything with direct argument values -- no sample frame yet.

> The download chunks below are shown but **not executed** in this vignette;
> they require network access to the NCCS bucket. Run them in a live session.

## Table abbreviations

The efile data is split into ~125 tables, each corresponding to a section of the
990. `panel990` ships short aliases for the most common ones. Preview them with
`table_catalog()`:

```{r}
table_catalog()
```

- **alias** -- the short token you pass to `tables=` (e.g. `"P08"`).
- **table** -- the canonical file name it resolves to.
- **cardinality** -- `1x1` is one row per filing (safe to merge side-by-side);
  `1xm` is one-to-many (kept separate unless you opt in).

You are not limited to the aliases: any literal table name (for a schedule or a
newly published table) is accepted as-is. `resolve_tables()` shows how a request
resolves:

```{r}
resolve_tables(c("P00", "P08", "SB-P01-T00-CONTRIBUTORS"))
```

## The data source

`data_source()` configures where files come from. The default points at the
NCCS bucket, so you rarely change it; pass a local directory to read from disk.

```{r}
src <- data_source()
src$root
```

## Downloading two tables for two years

`download_tables()` fetches (or reuses) the CSVs. Give it the years and tables
directly:

```{r, eval = FALSE}
dl <- download_tables(
  years  = 2019:2020,
  tables = c("P00", "P08"),     # header + revenue
  cache  = "retain",            # keep files in `path`; use "temporary" to discard
  path   = "efdata"
)
dl$manifest   # one row per table-year: status, bytes, path
```

Every table-year is recorded in a manifest, so you can see exactly what was
downloaded, reused, or failed.

## Read, merge, and stack

Reading turns the CSVs into data frames; merging joins the tables *within* a
year on the filing keys; stacking binds the years together.

```{r, eval = FALSE}
rd <- read_tables(dl)                 # optionally columns=, filters=
mg <- merge_tables(rd)                # join P00 + P08 per year by filing keys
```

`merge_tables()` joins the `1x1` tables side-by-side (header fields next to
revenue fields, one row per filing) and reports each join in a manifest. The
years are then stacked into a single data frame.

### The one-shot form

`panelize()` runs download -> read -> merge -> stack in a single call and
returns a `panel` object. The filing keys (`EIN2` / `TAX_YEAR` / `OBJECTID`)
are assigned automatically:

```{r, eval = FALSE}
p <- panelize(tables = c("P00", "P08"), years = 2019:2020)
p
#> <panel>  ... rows x ... cols  years 2019-2020
#>   sample frame: panel[P00,P08 | 2019-2020]  (0 rules, ... log entries)
#>   panel labels: stale / not computed
df <- as.data.frame(p)
```

Two tables, two years: the package downloads four files, merges the two tables
in each year, and stacks 2019 on top of 2020.

## Adding BMF fields, then filtering by state

Geography, NTEE, and organization type are **not** on the 990 itself -- they
live in the Business Master File (BMF). To filter by state you first join the
BMF, then subset. `bmf_merge()` downloads and attaches the native BMF fields by
`EIN2`:

```{r, eval = FALSE}
df <- bmf_merge(df)                   # appends geo_state_abbr, subsection_code, ntee_*, ...
bmf_vars()                            # the fields it adds by default
```

Now the state column exists, so a plain base-R subset works:

```{r, eval = FALSE}
ga <- df[df$geo_state_abbr == "GA", ]
```

That is the whole manual pipeline: pick tables and years, download, merge,
stack, attach the BMF, and filter. The next tutorial shows how a **sample
frame** captures these same choices as a reusable, self-documenting object.
