---
title: "Getting started with panel990"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with panel990}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(panel990)
```

`panel990` assembles IRS 990 efile tables into research-ready panels and gives
you a small vocabulary of verbs for describing, selecting, and cleaning them.
This tour walks the whole pipeline in six steps:

1. **Build** the panel from the raw efile tables
2. **Identify** each organization's membership pattern
3. **Filter** to the patterns your design calls for (including a balanced panel)
4. **Impute** missing interior years
5. **Smooth** noisy series
6. **Reconcile** the accounting identities that cleaning disturbs

Each step below pairs a short code example with a picture of what the verb does.
The download step needs network access, so it is shown but not run; every other
example runs on a small in-memory panel.

## 1. Build the panel

`panelize()` is the one-shot assembler: it downloads the tables you name for
the years you name, merges them within each year, stacks the years, and
(optionally) attaches organization traits from the Business Master File:

```{r, eval = FALSE}
panel <- panelize(
  tables = c("P00", "P01", "P08", "P09", "P10"),   # header, summary, revenue,
  years  = 2019:2022,                              #   expenses, balance sheet
  bmf    = TRUE
)
```

Within each year the five tables are joined into one wide row per filing. Every
table carries the same filing keys (`EIN2` × `TAX_YEAR` × `OBJECTID`), and the
join is a full outer join, so a filing missing from one part keeps its row:

![Five core 990 tables joined on the filing keys into one wide row per filing](figures/merge-tables.svg){width=100%}

The per-year tables share a harmonized layout, so they stack into one long
panel with one row per organization-year:

![Merged per-year tables stacked into one long panel with one row per organization-year](figures/stack-years.svg){width=100%}

`bmf = TRUE` then left-joins the BMF on `EIN2`, appending time-invariant
organization traits (name, NTEE, subsection, geography) to the right of the 990
fields and broadcasting them to every year the organization appears:

![BMF organization traits appended to the right of the 990 panel by a left join on EIN2](figures/bmf-merge.svg){width=100%}

Everything after this point works the same on a `panel` object or on any plain
data frame with an id and a year column, so the rest of the tour uses a small
in-memory panel:

```{r}
panel <- data.frame(
  EIN2     = c(rep("A", 5), rep("B", 3), rep("C", 3), rep("D", 2)),
  TAX_YEAR = c(2018:2022,   2020:2022,   2018:2020,   c(2018, 2022)),
  revenue  = c(10:14,       20:22,       30:32,       c(40, 44))
)
```

## 2. Identify panel types

Organizations rarely appear in every year of a panel. `panel_describe()`
classifies each one on two independent axes:

- **`panel_type`** (boundary) — which edges of the window the org touches:
  `persistent`, `entrant`, `exit`, `transient`.
- **`panel_spell`** (continuity) — whether its observed years are consecutive
  (`seamless`) or contain interior gaps (`segmented`).

![The eight panel membership patterns: panel_type crossed with panel_spell](figures/panel-types.svg){width=100%}

```{r}
panel_describe(panel, time = "TAX_YEAR", id = "EIN2")
```

`A` spans the window with no gaps (`persistent` + `seamless` — the balanced
case), `B` enters late, `C` exits early, and `D` spans the window but skips the
middle (`persistent` + `segmented`).

## 3. Filter panel types

`panel_filter()` keeps the rows of the organizations whose classification
matches. It classifies internally, so you just name what you want:

![panel_filter() selecting persistent organizations and entrants out of a mixed panel](figures/panel-filter.svg){width=100%}

```{r}
# organizations that span the whole window
unique(panel_filter(panel, panel_type = "persistent",
                    time = "TAX_YEAR", id = "EIN2")$EIN2)

# new entrants
unique(panel_filter(panel, panel_type = "entrant",
                    time = "TAX_YEAR", id = "EIN2")$EIN2)
```

### Searching for a balanced panel

A **balanced** panel is a rectangle: every retained organization observed in
every retained year. You can select it by pattern —
`panel_filter(panel_type = "persistent", spell = "seamless")` — or let
`panel_balance()` *search* for the year window that maximizes retained
observations:

![panel_balance() trimming a mixed panel to its largest fully observed block of organizations and years](figures/panel-balance.svg){width=100%}

```{r}
best <- panel_balance(panel, strategy = "max_rectangle",
                      id = "EIN2", time = "TAX_YEAR")
attr(best, "balance")
```

## 4. Impute missing cases

`panel_impute()` inserts an organization's missing interior years and fills
their numeric fields from the observations bracketing each gap
(`method = "interpolate"`, `"mean"`, `"locf"`, or `"nocb"`). Inserted rows are
flagged `imputed_row = TRUE` so they are never mistaken for filings:

![panel_impute() filling interior gaps, shown as a table with missing cells and as trend plots with imputed points](figures/panel-impute.svg){width=100%}

```{r}
filled <- panel_impute(panel, vars = "revenue", method = "interpolate")
filled[filled$EIN2 == "D", c("EIN2", "TAX_YEAR", "revenue", "imputed_row")]
```

`D`'s 2019–2021 hole is interpolated between its 2018 and 2022 filings. (The
researcher-facing wrapper `panel_complete()` fills every segmented span in one
call.)

## 5. Smooth values

`panel_smooth()` replaces a numeric field with a rolling weighted average taken
within each organization. `window` sets how many years are averaged; `weights`
sets how much the focal year keeps (`"equal"`, `"half"`, or `"decay"`):

![panel_smooth() applied with different window and weights settings, compared against the raw series](figures/panel-smooth.svg){width=100%}

```{r}
noisy <- data.frame(
  EIN2     = "A",
  TAX_YEAR = 2016:2022,
  revenue  = c(10, 12, 11, 50, 13, 12, 11)   # one-year spike
)
sm <- function(w, wt) round(panel_smooth(noisy, "revenue", window = w,
                                         weights = wt, verbose = FALSE)$revenue, 1)
data.frame(year = noisy$TAX_YEAR, raw = noisy$revenue,
           w3_equal = sm(3, "equal"), w5_equal = sm(5, "equal"),
           w3_decay = sm(3, "decay"))
```

## 6. Reconcile accounting principles

990 financials carry deterministic identities — contributions components sum to
the contributions total, revenue lines sum to total revenue, assets equal
liabilities plus net assets. Cell-level cleaning (partial imputation, rounding,
smoothing over filer errors) can leave rows that violate them.
`accounting_check()` finds the violations and `reconcile()` repairs them with
the smallest possible adjustment, holding the values you trust fixed:

![Independently smoothed revenue components no longer sum to total revenue; reconcile() restores the identity](figures/panel-reconcile.svg){width=100%}

```{r}
row <- data.frame(
  EIN2 = "A", TAX_YEAR = 2020,
  F9_08_REV_CONTR_FED_CAMP = 100, F9_08_REV_CONTR_MEMBSHIP_DUE = 50,
  F9_08_REV_CONTR_FUNDR_EVNT = 30, F9_08_REV_CONTR_RLTD_ORG = 20,
  F9_08_REV_CONTR_GOVT_GRANT = 180,   # imputed; the org reported a 500 total
  F9_08_REV_CONTR_OTH = 90,           # imputed
  F9_08_REV_CONTR_TOT = 500
)
accounting_check(row)                 # contributions short by 30

rec <- reconcile(row, fixed = setdiff(names(row), c("F9_08_REV_CONTR_GOVT_GRANT",
                                                    "F9_08_REV_CONTR_OTH")))
rec[, c("F9_08_REV_CONTR_GOVT_GRANT", "F9_08_REV_CONTR_OTH", "F9_08_REV_CONTR_TOT")]
accounting_check(rec)                 # identities restored
```

## Where to next

Each step has a task-focused vignette with the full details:

- *Downloading and assembling efile tables* and *the sampling framework* —
  building panels reproducibly
- *Panels and panel slices* — the classification vocabulary
- *Imputing missing years*, *Completing panel spans*, *Smoothing panel
  variables*, *Balancing a panel* — cleaning
- *Accounting consistency* and *Consistent gap-filling* — validation
