---
title: "Imputing missing years"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imputing missing years}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(panel990)
```

`panel_impute()` is the low-level engine that inserts a panel's missing
*interior* years and fills their numeric fields from the observations that
bracket each gap. (For the researcher-facing "just complete every span" verb,
see `panel_complete()` in *Completing panel spans*.)

A year can only be inserted if it is part of the panel, so our toy data has a
filler organization `F` observed every year, and one organization `A` that
skips the middle:

```{r}
df <- data.frame(
  EIN2     = c("A", "A", "F", "F", "F", "F"),
  TAX_YEAR = c(2018, 2021, 2018, 2019, 2020, 2021),
  revenue  = c(0, 30, 5, 5, 5, 5)
)
```

`A` spans 2018--2021 but is missing 2019 and 2020 -- `persistent` on the
boundary axis, `segmented` on the spell axis:

```{r}
attr(panel_describe(df, print = FALSE), "classification")[, c("EIN2", "panel_type", "panel_spell")]
```

## Fill methods

The gap between `A`'s 2018 value (0) and 2021 value (30) can be filled four
ways. Compare them on the two missing years:

```{r}
fill <- function(m) {
  out <- panel_impute(df, vars = "revenue", method = m)
  out$revenue[out$EIN2 == "A" & out$TAX_YEAR %in% c(2019, 2020)]
}
data.frame(
  year        = c(2019, 2020),
  mean        = fill("mean"),        # flat average of the brackets
  interpolate = fill("interpolate"), # linear between the brackets
  locf        = fill("locf"),        # last observation carried forward
  nocb        = fill("nocb")         # next observation carried backward
)
```

`mean` (the default here) puts the same average in every hole; `interpolate`
draws a straight line; `locf` holds the prior value; `nocb` pulls the next value
back. Imputation never extends past an organization's own span, and a hole stays
`NA` only when *both* brackets are missing.

## Which organizations are eligible

By default only `persistent` organizations (those that span the window) are
imputed, and every gap is filled:

```{r}
out <- panel_impute(df, vars = "revenue")     # method = "mean", types = "persistent"
out[out$EIN2 == "A", c("TAX_YEAR", "revenue", "imputed_row")]
```

Inserted rows are flagged with `imputed_row`, so you can always tell filled
values from reported ones. Gap-size and gap-count limits skip anything too wide
to trust:

```{r}
# refuse to fill a gap longer than one year
nrow(panel_impute(df, vars = "revenue", max_gap_size = 1))   # A's 2-year gap is left alone
nrow(df)
```

Other controls: `vars` chooses which numeric fields to fill (all numeric
non-key fields by default), `types` widens eligibility to other boundary types,
and `as_integers` rounds filled values for originally integer columns.
