---
title: "3. Panels and panel slices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Panels and panel slices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(panel990)
```

With several years assembled, `panel990` classifies each organization by *how
its observations sit in the panel window*, so you can slice out entrants, exits,
or a balanced set.

In a real project you would start from a download -- for example
`panelize(tables = c("P00", "P01"), years = 2016:2022, bmf = TRUE)` -- but the classification
works on any data frame with an id and a year, so this tutorial uses a small
in-memory panel.

```{r}
panel <- data.frame(
  EIN2     = c(rep("A", 5), rep("B", 3), rep("C", 3), rep("D", 2)),
  TAX_YEAR = c(2018:2022,   2020:2022,   2018:2020,   c(2018, 2022)),
  revenue  = c(10:14,       20:22,       30:32,       c(40, 44)),
  stringsAsFactors = FALSE
)
```

## Describe the panel

Classification has **two independent axes**:

- **`panel_type`** (boundary): `persistent` (present at the first *and* last
  year), `entrant` (enters late, present at the end), `exit` (present at the
  start, gone before the end), `transient` (only interior years), `empty`.
- **`panel_spell`** (continuity): `seamless` (consecutive years) or `segmented`
  (interior gaps).

```{r}
panel_describe(panel, time = "TAX_YEAR", id = "EIN2")
```

Organization A spans 2018--2022 with no gaps (`persistent` + `seamless` -- the
"balanced" case); D also spans the window but skips the middle years
(`persistent` + `segmented`); B enters in 2020 (`entrant`); C leaves after 2020
(`exit`).

## Slice by membership

`panel_filter()` returns the rows of the organizations that match. It
classifies internally, so you just name what you want:

```{r}
# new entrants
unique(panel_filter(panel, panel_type = "entrant",
                    time = "TAX_YEAR", id = "EIN2")$EIN2)

# exits
unique(panel_filter(panel, panel_type = "exit",
                    time = "TAX_YEAR", id = "EIN2")$EIN2)

# a balanced panel = spans the window with no gaps
unique(panel_filter(panel, panel_type = "persistent", spell = "seamless",
                    time = "TAX_YEAR", id = "EIN2")$EIN2)
```

`min_obs=` adds a minimum number of observed years, and passing a
pre-computed classification (from `panel_describe()`) avoids reclassifying.

## Attach labels for your own analysis

`panel_label()` appends the classification columns to every row, preserving
order -- handy for grouping or modelling:

```{r}
labelled <- panel_label(panel, time = "TAX_YEAR", id = "EIN2")
labelled[!duplicated(labelled$EIN2),
         c("EIN2", "panel_type", "panel_spell", "panel_year_count")]
```

## Slicing through a sample frame

Inside a sample frame, `classify_panel()` stores the two axes as label rules, so
panel membership becomes just another filter you can compose with the others:

```{r}
sfw <- classify_panel(create_sfw("panel demo"), panel)
get_rules(sfw)

# keep only the window-spanning organizations
persistent <- apply_sfw(panel, sfw, panel_type = "persistent", verbose = FALSE)
unique(persistent$EIN2)
```

Combined with the filters and BMF traits from the previous tutorial, this lets a
single frame express a full study sample -- "Georgia 501(c)(3) housing
nonprofits observed as a balanced panel from 2019 to 2021" -- and apply it
consistently across every step of the analysis.
