---
title: "5. Private foundations (990PF)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Private foundations (990PF)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(panel990)
```

Private foundations file Form 990PF rather than the Form 990 or 990EZ. NCCS
publishes their efile data as a separate release with its own tables, and
`panel990` reads it through the same machinery as the 990 release. This
tutorial covers what is different.

> The download chunks below are shown but **not executed** in this vignette;
> they require network access to the NCCS bucket. Run them in a live session.

## One release per source

Each NCCS release is published as two databases:

| Form family | Filers | S3 prefix |
|---|---|---|
| `"990"` | Form 990 and 990EZ, told apart by `RETURN_TYPE` | `efile_` |
| `"990PF"` | Form 990PF | `efilepf_` |

A `data_source()` reads exactly one of them, chosen with `form`:

```{r}
data_source(form = "990PF")$root
```

A panel never mixes the two. The 990PF financial statements are a different
structure from the 990 parts, not a relabelled copy of them, so there is no
single set of columns on which a foundation and a public charity line up. Build
990 and 990PF panels separately.

## 990PF tables

The 990PF release has its own catalog. Its parts get their own `PF` aliases, so
no short name means one table in a 990 source and another in a 990PF source:

```{r}
pf <- table_catalog(form = "990PF")
pf[!is.na(pf$alias), ]
```

`P00` is the one alias the two releases share, because both publish the
filing header under the same name, `F9-P00-T00-HEADER`. The 990PF release also
carries its own copies of the signature and Schedule B tables.

Asking for a table from the other release is an error, raised before any
download starts:

```{r, error = TRUE}
resolve_tables("PF01", data_source())          # a 990 source
resolve_tables("P08", data_source(form = "990PF"))
```

## Building a panel

`panelize_pf()` is `panelize()` with a 990PF source. It defaults to the 990PF
header and the Part I and Part II financial statements:

```{r, eval = FALSE}
pf <- panelize_pf(years = 2021:2022, backend = "duckdb", cache = "none")
```

Everything else works as it does for a 990 panel: sample frames, the BMF
join, deduplication, and the provenance log. Retained downloads are cached under
`<path>/990PF/<year>/`, because the 990PF release reuses the shared table
names and its files must not overwrite the 990 release's.

## Reading the 990PF financial statements

Part I (`PF01`) reports each line in up to four columns, marked by the suffix:

| Suffix | Column |
|---|---|
| `_BOOKS` | (a) revenue and expenses per books |
| `_NET` | (b) net investment income |
| `_ADJ_NET` | (c) adjusted net income |
| `_DISBMT` | (d) disbursements for charitable purposes |

Part II (`PF02`) reports the balance sheet at beginning- and end-of-year book
value (`_BOY_BV`, `_EOY_BV`) and end-of-year fair market value (`_EOY_FMV`).
Part III (`PF03`) reconciles net assets from the beginning to the end of the
year.

The bundled `field_concordance_pf` describes every 990PF field, and
`financial_fields(form = "990PF")` lists the money fields in Parts I-III:

```{r}
head(financial_fields(form = "990PF"))
```

## Blanks and form scope

A 990PF return is one form, so form scope, which separates full-990 from
990EZ fields, does not apply. A sample frame that selects columns by a 990
scope such as `"both"` is an error in a 990PF panel; select by `tables` or
`vars` instead.

Blanks are common in the 990PF financial statements, and `panel_normalize()`
reads them the same way it does for the 990: a blank money field on a filed
return is zero, while a row with no financial data at all is left untouched.
The same call works for either kind of panel:

```{r, eval = FALSE}
pf <- panel_normalize(pf)
```

## Accounting identities

`accounting_identities` includes 29 990PF identities, marked
`form_scope == "PF"`. They cover the Part I column totals, the Part II
balance-sheet equation and subtotals, and the Part III reconciliation and its
ties to Parts I and II. `accounting_check()` evaluates whichever identities
have all their fields in the data, so the call is the same as for a 990 panel:

```{r, eval = FALSE}
accounting_check(panel_data(pf), section = "balance_sheet")
```

```{r}
ids <- accounting_identities[accounting_identities$form_scope == "PF", ]
unique(ids[, c("identity", "section", "description")])[1:6, ]
```
