---
title: "2. The sampling framework"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. The sampling framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(panel990)
```

The first tutorial specified tables, years, and filters by hand. A **sample
frame** (`sfw`) captures those same choices as a single reusable object -- a
registry of *keys* and typed *rules* that you can apply to any data frame,
check against, and hand to `panelize()` to govern a build.

## Build a frame

A frame is just a set of rules. Friendly arguments to `create_sfw()` are
convenience filters that lower into rules:

```{r}
sfw <- create_sfw(
  "housing 2024",
  state       = "GA",     # -> filter on geo_state_abbr
  filter_501c = "3",      # -> filter on subsection_code (501(c)(3))
  years       = 2019:2021 # -> filter on TAX_YEAR
)
sfw
```

Entity (`EIN2`) and time (`TAX_YEAR`) keys are registered by default. Add the
filing-level key so deduplication and joins know a unique record:

```{r}
sfw <- add_key(sfw, "filing", "unique_record", "OBJECTID")
get_keys(sfw)
```

## Structured rules

Beyond the sugar, `add_rule()` adds any rule directly. Filters are
**structured** -- a column, an operator, and values -- which keeps them
inspectable and portable:

```{r}
sfw <- add_rule(sfw, "revenue positive", "filter",
                column = "F9_01_REV_TOT_CY", op = ">", values = 0)
get_rules(sfw)
```

`op` may be `in`, `not_in`, `==`, `!=`, `>`, `>=`, `<`, `<=`, `between`,
`is_true`, `is_false`; for an arbitrary predicate use `expr = "..."`.

Most rules are filters: each one names a column, an operator, and the values it
tolerates — which is to say, each one names exactly the rows it removes:

![Four filter rules drawn as colored strips over a tall data frame, each strip covering the rows that rule removes](figures/sfw-rules.svg){width=100%}

## Apply the frame

`apply_sfw()` runs the rules on a data frame. Here is a tiny stand-in for a
BMF-merged panel:

```{r}
df <- data.frame(
  EIN2             = c("EIN-11-1111111", "EIN-22-2222222", "EIN-33-3333333"),
  TAX_YEAR         = 2020,
  geo_state_abbr   = c("GA", "GA", "FL"),
  subsection_code  = c("3", "4", "3"),
  F9_01_REV_TOT_CY = c(500000, 0, 250000),
  stringsAsFactors = FALSE
)

out <- apply_sfw(df, sfw, verbose = FALSE)
out
```

Only the Georgia 501(c)(3) with positive revenue survives. Every step is
recorded as a manifest attribute -- a reproducible record of what each rule
removed:

```{r}
attr(out, "sfw_steps")
```

The manifest is the frame's audit trail: one row per rule, recording how many
rows each one removed on its way to the final panel:

![The sfw_steps manifest as a ledger: each rule's rows in, rows removed, and rows out, with a shrinking bar showing what dropped](figures/sfw-manifest.svg){width=100%}

`conform()` answers the inverse question -- does a data frame already satisfy
the frame? -- without changing it:

```{r}
conform(df, sfw, verbose = FALSE)$rows_violating
```

## Prefiltering before merges makes it faster

When you hand the frame to `panelize()`, it governs the build:

```{r, eval = FALSE}
p <- panelize(sfw = sfw, tables = c("P00", "P01", "P08"), years = 2019:2021, bmf = TRUE)
```

Two restrictions are applied **before** the expensive table merges, so the
package does far less work on a large panel:

- **Years** -- only the requested years' files are downloaded.
- **Entities** -- if the frame restricts to a set of organizations, that set is
  pushed down to *read time*. Because `EIN2` is present in every efile table,
  only those organizations' rows are ever read, and only they are carried into
  the merges.

Trait filters (state, 501(c) type, NTEE) live in the BMF, so they resolve
*after* the BMF merge. To turn a trait restriction into a read-time prefilter,
resolve it to an organization set from the BMF first, then add it as a `subset`
rule:

```{r, eval = FALSE}
bmf   <- bmf_retrieve(states = "GA")                      # just the GA mart
ga_c3 <- bmf$EIN2[bmf$geo_state_abbr == "GA" &
                    bmf$subsection_code == "3"]

sfw <- add_rule(sfw, "GA 501c3 cohort", "subset", subset = ga_c3)
p   <- panelize(sfw = sfw, tables = c("P00", "P01", "P08"), years = 2019:2021)
```

![Two large tables prefiltered down to slim cohort tables before the merge; the full-size merge risks exhausting RAM](figures/sfw-prefilter.svg){width=100%}

Now `panelize()` reads only those organizations before merging tables and
stacking years -- the trait filter has become a fast pushdown. The remaining
rules still apply after assembly as the source of truth, and the whole pipeline
is documented in the resulting panel's step manifest.
