---
title: "Smoothing panel variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Smoothing panel variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(panel990)
```

`panel_smooth()` replaces a numeric field with a rolling weighted average taken
*within each organization*, ordered by year. It dampens year-to-year noise
(reporting quirks, one-off spikes) while leaving the rows in place.

A single organization with a one-year spike:

```{r}
df <- data.frame(
  EIN2     = "A",
  TAX_YEAR = 2016:2022,
  revenue  = c(10, 12, 11, 50, 13, 12, 11)   # spike in 2019
)
```

## Window and weights

`window` is the (odd) number of years averaged; `weights` decides how much each
year in the window counts:

- **`equal`** -- a plain moving average; every year in the window counts the same.
- **`half`** -- the focal year gets half the weight, the neighbours share the other half.
- **`decay`** -- weight falls by half per year of distance (`0.5^distance`).

```{r}
w <- function(wt) panel_smooth(df, "revenue", window = 3, weights = wt,
                               verbose = FALSE)$revenue
data.frame(
  year    = df$TAX_YEAR,
  raw     = df$revenue,
  equal   = round(w("equal"), 1),
  half    = round(w("half"), 1),
  decay   = round(w("decay"), 1)
)
```

Notice 2019: `equal` pulls the spike down hardest (it treats the neighbours and
the spike alike), while `half` and `decay` keep more of the focal value. (For a
centred width-3 window `half` and `decay` give identical weights; they differ at
the series edges, where the window is off-centre, and more visibly from
`window = 5` on, where `decay` down-weights distant years more steeply.)

A wider window smooths more aggressively:

```{r}
data.frame(
  year = df$TAX_YEAR,
  w3   = round(panel_smooth(df, "revenue", window = 3, verbose = FALSE)$revenue, 1),
  w5   = round(panel_smooth(df, "revenue", window = 5, verbose = FALSE)$revenue, 1)
)
```

## Edges, gaps, and multiple series

At the ends of a series the fixed-width window slides inward rather than
shrinking, and organizations with fewer years than `window` use a smaller
effective window. Missing values (`NA`/`NaN`) are dropped from each window and
the remaining weights are renormalized, so a stray gap does not blank out its
neighbours; a window with nothing observed returns `NA`.

`panel_smooth()` handles many organizations and several fields at once, always
smoothing each organization independently and returning the rows in their
original order:

```{r}
panel <- data.frame(
  EIN2     = rep(c("A", "B"), each = 4),
  TAX_YEAR = rep(2019:2022, 2),
  revenue  = c(100, 300, 200, 400,  50, 40, 90, 60),
  expenses = c(90,  250, 180, 350,  45, 35, 80, 55)
)
head(panel_smooth(panel, vars = c("revenue", "expenses"),
                  window = 3, verbose = FALSE), 4)
```

Smoothing changes values but never the shape of the data -- same rows, same
order, selected columns replaced by their smoothed versions.
