---
title: "3. Cross-year requirements on the source"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Cross-year requirements on the source}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(panel990)
```

Every rule in the previous tutorial is a statement about rows you already
have. `state = "GA"` can be checked one row at a time; so can
`filter_501c = "3"`. `apply_sfw()` evaluates them against a data frame and
that is the end of it.

Some sampling criteria are not like that:

> Which organizations filed in **all three** of 2019, 2020 and 2021?

> Which were **990EZ filers every year**, never switching to the full 990?

Neither question can be answered from a single row, or even from a single
table-year. They are facts about the *source*. Worse, the obvious way to
answer them — build the panel, then filter it — reads exactly the data you
were hoping to avoid reading. You download three years of every column to
discover that you wanted 5,560 organizations out of 241,552.

A `require` rule states such a condition, and `resolve_frame()` answers it
against the source before the real read happens.

## Declaring a requirement

`require` is a rule type like any other, so it goes on the frame the same way:

```{r}
frame <- create_sfw("balanced 990EZ filers", record = "OBJECTID")
frame <- add_rule(frame, "balanced", type = "require", present_in = "all")
frame <- add_rule(frame, "always_ez", type = "require", present_in = "any",
                  column = "RETURN_TYPE", op = "in", values = "990EZ",
                  holds = "every")
get_rules(frame)
```

Two arguments carry the cross-year meaning:

`present_in` — how many of the requested years the organization must appear in.
`"all"` (the default) gives a balanced frame, `"any"` imposes no presence
requirement, and an integer sets a minimum: `present_in = 2` accepts anything
observed in at least two years.

`holds` — how a per-filing predicate folds over the years. `"every"` (the
default) requires `RETURN_TYPE == "990EZ"` in *every* year observed; `"any"`
requires it in at least one. The predicate itself is the ordinary structured
`column`/`op`/`values` triple used by `filter` rules.

The two combine, so `present_in` and `holds` are independent choices:

| `present_in` | `holds` | Reads as |
|---|---|---|
| `"all"` | `"every"` | in every year, and a 990EZ filer in each |
| `"any"` | `"every"` | a 990EZ filer in every year it *did* file |
| `"all"` | `"any"` | in every year, and a 990EZ filer at least once |

An `expr` string is refused rather than quietly ignored, because an opaque
predicate cannot be pushed to the source:

```{r, error = TRUE}
add_rule(frame, type = "require", expr = "RETURN_TYPE == '990EZ'")
```

## Resolving against the source

`resolve_frame()` executes the requirements and rewrites the frame:

```{r, eval = FALSE}
frame <- resolve_frame(frame, years = 2009:2011,
                       source = data_source(format = "parquet"))
#> <- REQUIRE  balanced [F9-P00-T00-HEADER x 3 year(s)]: 33,946 EIN2 in 0.04s
#> <- REQUIRE  always_ez [F9-P00-T00-HEADER x 3 year(s)]: 94,422 EIN2 in 0.06s
#> resolve_frame: 5,560 EIN2 satisfy 2 requirement(s).
```

Multiple requirements intersect: an organization must satisfy all of them.
33,946 organizations filed in all three years, and 94,422 never filed anything
other than a 990EZ — 5,560 satisfy both.

What comes back is an ordinary frame whose sample is now concrete:

```{r, eval = FALSE}
get_rules(frame)
#>        name    type active                                               detail
#> 1  balanced require  FALSE                                  P00: present in all
#> 2 always_ez require  FALSE P00: present in any; every year RETURN_TYPE in 990EZ
#> 3  resolved  subset   TRUE                                             5560 ids
```

The requirements are **deactivated, not dropped**, and the answer arrives as a
captured `subset` rule. That is the whole design:

- The frame stays declarative and source-agnostic. `apply_sfw()` never learns
  about years or files — it only ever sees a `subset`, which it already knows
  how to execute against any data frame.
- The sample is reproducible and auditable. The ids are captured by value, and
  the frame still records the *question* next to the answer.

`manifest()` logs the resolution, so a panel built this way explains where its
sample came from:

```{r, eval = FALSE}
manifest(frame)
#>                step                                               detail rows_after
#> 1  require: balanced                                  P00: present in all      33946
#> 2 require: always_ez P00: present in any; every year RETURN_TYPE in 990EZ      94422
#> 3      resolve_frame                         2 requirement(s) intersected       5560
```

## Why it is cheap

Resolution reads the entity key, the time key, and any column a predicate
names — nothing else. The financial columns are never touched, because the
question does not involve them.

On a parquet release that is two or three columns of one table. Answering
"who filed in all three years" over the 2009–2011 header table reads **1.4 MB**
of `EIN2` and `TAX_YEAR`. The same three files as CSV are **273 MB**, and a CSV
scan has to read all of it because a text file has no column boundaries to
seek to.

```{r, eval = FALSE}
# Both resolve to the same 5,560 ids.
data_source(format = "parquet")   # 0.17s
data_source(format = "csv")       # 1.10s locally, far worse over the network
```

The requirement is answered the same way on either format — `resolve_frame()`
works on a CSV release too. Parquet just makes it nearly free, and the gap
widens without a local cache, where column pruning becomes a handful of HTTP
range requests instead of a whole-file transfer. See `data_source()` for the
format's type contract.

## Handing it to panelize()

`panelize()` resolves an outstanding requirement on its own, before the read
it is meant to narrow:

```{r, eval = FALSE}
panel <- panelize(
  frame, tables = c("P00", "P01"), years = 2009:2011,
  source  = data_source(format = "parquet"),
  backend = "duckdb",       # pushes the subset into the scan
  cache   = "none",         # read the published files in place
  unique_rows = FALSE       # let the column projection reach the scan
)
```

There is no need to call `resolve_frame()` first — the frame above would work
passed straight to `panelize()`. Resolving it by hand is useful when you want
to inspect the sample, reuse one id set across several builds, or pin it for
reproducibility.

The three acquisition arguments are what turn the resolved subset into a
narrow read rather than a filter after the fact. `unique_rows = FALSE`
matters more than it looks: deduplication compiles to `SELECT DISTINCT *`,
which needs every column and so blocks the projection from reaching the scan.
The efile build emits no exact duplicate rows, so switching it off is both
faster and equivalent.

## Coverage is not one row per year

A requirement guarantees that an organization *appears* in each year. It does
not guarantee one row per year, because an amended return is a second filing
for the same tax year:

```{r, eval = FALSE}
table(table(panel_data(panel)$EIN2))
#>    3    4    5
#> 5508   49    3
```

That is a different job, and there is already a rule for it. Add a `dedup`
rule and each organization-year collapses to one filing:

```{r, eval = FALSE}
panel <- panelize(add_rule(frame, type = "dedup"), tables = c("P00", "P01"),
                  years = 2009:2011, source = data_source(format = "parquet"),
                  backend = "duckdb", cache = "none", unique_rows = FALSE)
table(table(panel_data(panel)$EIN2))
#>    3
#> 5560
```

Keep the two concerns separate: `require` decides *who is in the sample*,
`dedup` decides *which filing represents an organization-year*. See
`deduplicate()` for how the winning filing is chosen.

## Choosing between require and filter

A `filter` rule is the right tool whenever the criterion can be read off one
row. Reach for `require` only when the condition spans years:

| Criterion | Rule |
|---|---|
| filed a 990EZ *this* year | `filter` on `RETURN_TYPE` |
| a 990EZ filer *every* year | `require`, `holds = "every"` |
| in Georgia | `filter` on `geo_state_abbr` (BMF) |
| present in all requested years | `require`, `present_in = "all"` |
| observed at least twice | `require`, `present_in = 2` |

`require` also overlaps with `panel_balance()`, and the difference is when the
work happens. `panel_balance()` balances a panel you have already built;
`require` resolves the balanced sample first and never reads the rows it is
going to drop. Use `panel_balance()` when the panel is already in memory, and
`require` when you are about to acquire it.
